The Jester joke recommender system

### 2020-08-03 inst/f1points.Rmd
minor tweaks

---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address:
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The points scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes being controversial.
  The inaugural points system, in use from 1950
  to 1959 was $(8,6,4,3,2)$: 8 points to the winner, 6 to second
  place, and so on.  In 2020 the points system was changed to
  $(25,18,15,12,10,8,6,4,2,1)$.  The overall ranking of the
  competitors is determined by summing points awarded over a season;
  but this system suffers from the defect that the points system
  itself is intrinsically arbitrary.  Here, I apply generalized
  Plackett-Luce likelihood functions to Formula 1 motor racing, and
  use maximum likelihood techniques to identify a ranking for the competitors;
  this method includes no use of an arbitrary points system.
  Using this methodology, I assess a number of reasonable points
  scoring systems objectively.  In a well-defined statistical sense, a
  ``winner-takes-all'' points system outperforms the other points
  systems considered.
keywords: 
  # at least one keyword must be supplied
  formatted: [Bradley-Terry, Plackett-Luce, order statistics, rank
    statistics, points systems, Formula 1]
  plain:     [Bradley-Terry, Plackett-Luce, order statistics, rank
    statistics, points systems, Formula 1]
preamble:
  \usepackage{amsmath}
  \usepackage{amssymb}
output: rticles::jss_article
---

@@

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points allocation
system wherein competitors are awarded points based on race finishing
order; points accumulate additively.  The overall competition winner
is the competitor who accumulates the most points after the final
race.

However, in the case of Formula 1 Motor Racing, the points system is
the subject of much controversy, having changed often since the
competition's inauguration in 1950 when the points allocation, used
from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only the first
five finishers.  As of 2020, the current system of
$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.  Still, it is reasonable to assume that each
driver strives to maximise his rank.

@@

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
Noting that there is information in the whole of the finishing order,
and not just the first across the line, we can follow
\cite{plackett1975} and consider the runner-up to be the winner among
the remaining competitors, and so on down the finishing order.
Without loss of generality, if the order of finishing were
$1,2,3,4,5$, then a suitable
\citeauthor{plackett1975}-\citeauthor{luce1959} likelihood function
would be

@@

We see in figure \ref{piechartstrength} that the driver with the
largest estimated strength was Rosberg in 2016 (at about 49\%) and in
years 2017-8-9 was Hamilton (at about 29\%, 35\%, and 50\%
respectively).  However, it should be noted that a likelihood ratio
test [`samep.test()`, supplied with the package] does not reject, for
example, the null that Hamilton and Vettel have the same strength in
2017 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
$e^{1.5}\simeq 4.48$, corresponding (by Wilks's theorem) to an
asymptotic $p$-value of about $0.08$.

```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
ignore <- samep.test(H2017,1:2)
```

## Likelihood scoring vs points scoring

@@

We can plot one ranking against the other\footnote{The historically
correct points awarded to the drivers differs from that calculated
here.  That is for two reasons: firstly, ``fastest lap'' points are
not included here, and also the truncation of the order table to the
first 11 drivers can increase the rank of a driver if non-first-11
drivers are placed}, shown in Figure
\ref{orderingbypointsandlikelihood}.  We thus see a comparison between
the two ordering systems.  Taking 2017 as an example, drivers Hamilton
and Vettel are are respectively first and second according to both
ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT IS
WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.  We
define the \emph{degree of agreement} between the two ranking systems
as the maximal value of $n$ such that places $1,2,\ldots n$ all match.
Thus, from figure \ref{orderingbypointsandlikelihood}, the degree of
agreement between likelihood ranking and points ranking for the years
2016-2019 would be 8,2,6, and 2 respectively.

@@

We now assess the six points systems using the methodology outlined
above, using 1000 _in silico_ trials.  For each of the six points
systems, each of the 1000 trials results in a single non-negative
integer: the degree of agreement between the Plackett-Luce ranking and
the rankings according to the points system considered.  There are
three measures that might be used to assess the distribution of degree
of agreement: (1), the mean degree of agreement $\overline{d}$; (2),
the probability of correctly predicting the winner,
$\operatorname{Pr}(d\geqslant 1)$; and (3), the probability of
correctly predicting the complete order statistic
$\operatorname{Pr}(D=n)$.  These summaries are shown for each of the
six points systems in figures \ref{simulated_mean},
\ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
respectively.  These figures differ in detail but all have one
feature: the winner-takes-all points system consistently outperforms
the other systems.  This is particularly evident in figure
\ref{simulated_complete_agreement}.

@@

Many competitive situations involve ranking competitors and one common
method for combining repeated observations is to assign points for the
winner, second place, third place, etc.  The overall ranking is
decided on the basis of accumulated points.  However, statistical
analysis made difficult by the arbitrary values of the points given.

The points system used in Formula 1 motor racing is a source of lively
debate from many perspectives, with changes being controversial.  
Here I treat total points scored as a random variable and compare
different point allocation schemes against objective Plackett-Luce
ranks.  By several criteria, a ``winner takes all'' performs better
than other reasonable points systems.  However, such a system
furnishes little incentive for mid-ranking competitors and as such
might result in duller races.

>>>>

---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address:
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The points scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes being controversial.
  The inaugural points system, in use from 1950
  to 1959 was $(8,6,4,3,2)$: 8 points to the winner, 6 to second
  place, and so on.  In 2020 the points system was changed to
  $(25,18,15,12,10,8,6,4,2,1)$.  The overall ranking of the
  competitors is determined by summing points awarded over a season;
  but this system suffers from the defect that the points system
  itself is intrinsically arbitrary: the points system does not affect
  the competitors' behaviour but can change the overall ranking.
  Here, I calculate generalized Plackett-Luce likelihood functions for
  Formula 1 motor racing, and use maximum likelihood techniques to
  identify a ranking for the competitors; this method includes no
  arbitrary points system. Using this methodology, I assess a number
  of reasonable points scoring systems  objectively,  by comparing
  likelihood-based rankings with points-based rankings.  In one
  well-defined statistical sense, a linear  points system outperforms
  the other points systems considered, and in another the ``winner
  takes all'' system is the best.
keywords: 
  # at least one keyword must be supplied
  formatted: [Bradley-Terry, Plackett-Luce, order statistics, rank
    statistics, points systems, Formula 1]
  plain:     [Bradley-Terry, Plackett-Luce, order statistics, rank
    statistics, points systems, Formula 1]
preamble:
  \usepackage{amsmath}
  \usepackage{amssymb}
output: rticles::jss_article
---

@@

Formula 1 racing is an important and prestigious motor sport
\citep{test}.  Season ranking is based on a points allocation
system wherein competitors are awarded points based on race finishing
order; points accumulate additively.  The overall competition winner
is the competitor who accumulates the most points after the final
race.

However, in the case of Formula 1 Motor Racing, the points system is
the subject of much controversy, having changed often since the
competition's inauguration in 1950 when the points allocation, used
from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only the first
five finishers.  As of 2020, the current points system of
$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.  Still, it is reasonable to assume that each
driver strives to maximise his rank and this will be done here.  If
this is so, then changing the points system might change the
competitors' rankings but not their behaviour: surely a defect of
using points to rank the competitors.

@@

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow \cite{plackett1975} and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order.  Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable \citeauthor{plackett1975}-\citeauthor{luce1959} likelihood
function would be

@@

We see in figure \ref{piechartstrength} that the driver with the
largest estimated strength was Rosberg in 2016 at about 30\%, and in
years 2017-8-9 was Hamilton (at about 29\%, 37\%, and 42\%
respectively).  As an illustration of the value of likelihood methods
(as opposed to points-based methods), a likelihood ratio test
[`samep.test()`, supplied with the package] rejects
the null that Hamilton and Vettel have the same strength in
2018 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
$e^{2.76}\simeq 15.8$, corresponding (by Wilks's theorem) to an
asymptotic $p$-value of about $0.02$.

```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
ignore <- samep.test(H2018,c("Hamilton","Vettel"))
```
## Likelihood scoring vs points scoring

@@

We can plot one ranking against the other\footnote{The historically
correct points awarded to the drivers differs from that calculated
here.  That is for two reasons: firstly, ``fastest lap'' points are
not included here, and also the truncation of the order table to the
first 11 drivers can increase the rank of a driver if non-first-11
drivers are placed}, shown in Figure
\ref{orderingbypointsandlikelihood}.  We thus see a comparison between
the two ordering systems.  Taking 2017 as an example, drivers Hamilton
and Vettel are are respectively first and second according to both
ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT IS
WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.  We
define the \emph{degree of agreement} between the two ranking systems
as the maximal value of $r$ such that places $1,2,\ldots r$ all match.
Thus, from figure \ref{orderingbypointsandlikelihood}, the degree of
agreement between likelihood ranking and points ranking for the years
2016-2019 would be 8,2,6, and 2 respectively.

@@

We now assess the six points systems using the methodology outlined
above, using 1000 _in silico_ trials.  For each of the six points
systems, each of the 1000 trials results in a single non-negative
integer: the degree of agreement between the Plackett-Luce ranking and
the rankings according to the points system considered.  There are
three measures that might be used to assess the distribution of degree
of agreement: (1), the mean degree of agreement $\overline{d}$; (2),
the probability of correctly predicting the winner,
$\operatorname{Pr}(d\geqslant 1)$; and (3), the probability of
correctly predicting the complete order statistic
$\operatorname{Pr}(D=n)$.  These summaries are shown for each of the
six points systems in figures \ref{simulated_mean},
\ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
respectively.

We see that the linear points system $(n,n-1,n-2,\ldots,3,2,1)$ gives
the highest mean number of places, and also the highest probability of
correctly predicting the winner.  However, the probability of
predicting the complete order statistic is maximized using the winnner
takes all system $(1,0,0,\ldots)$.

@@

Many competitive situations involve a ranking of the participants and
one common method for combining repeated observations is to assign
points for the winner, second place, third place, etc.  The overall
ranking is decided on the basis of accumulated points.  However,
points systems are unavoidably arbitrary.  Assuming that participants
attempt to optimize their placing means that the arbitrary
nature of any points system.

Maximum likelihood estimation of participants' Plackett-Luce strengths
furnishes a ranking system that does not suffer from the arbitrariness
of a points system.

The points system used in Formula 1 motor racing is a source of lively
debate from many perspectives, with changes being controversial.  Here
I treat total points scored as a random variable and compare different
point allocation schemes against objective Plackett-Luce ranks.  Of
the six points systems considered here, a linear system or a
winner-takes-all-system appear to be closest to objective
Plackett-Luce, depending on the exact definition of ``closest''.
